---
title:
author: Per Unneberg
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  revealjs::revealjs_presentation:
    css: reveal.css
    includes:
      in_header: footer.html
    self_contained: true
    highlight: tango
    fig_width: 10
    fig_height: 8
    fig_caption: false
    toc: true
    toc_depth: 2
    slide_level: 1
    reveal_options:
      slideNumber: true
      previewLinks: true
      minScale: 1
      maxScale: 1
      height: 1400
      width: 1200
---

```{r  snakemake-byoc-2020-knitr, echo=FALSE, eval=TRUE, include=TRUE }
library(knitr)
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      fig.width=12, fig.height=10, autodep=TRUE, echo=TRUE,
                      cache=FALSE, include=TRUE, eval=FALSE, tidy=FALSE, error=TRUE,
                      class.source = "numberLines lineAnchors",
                      class.output = c("numberLines lineAnchors chunkout"))
knitr::knit_hooks$set(inline = function(x) {
                      prettyNum(x, big.mark=" ")
}) 
```

```{r  snakemake-byoc-2020-libs, echo=FALSE, eval=TRUE, cache=FALSE }
library(ggplot2)
library(viridis)
bw <- theme_bw(base_size=24) %+replace% theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
theme_set(bw) 
```

#

<div>
<section>
<h1 class="title">Snakemake BYOC 2020</h1>
<h2 class="subtitle">Snakemake profiles</h2>
<h2 class="author">Per Unneberg</h2>
<h3 class="date">2020-10-06</h3>
</section>
</div>


# Contents

<div style="padding:100px">
<h3>Cluster execution</h3>
<h3>Snakemake profiles</h3>
</div>

<div class="fragment" style="padding:100px">
<h3>Setup</h3>

-   Very simple examples with snakefiles and code to run
-   All snakefiles and code is available in code repository (show path)
-   NB: slurm examples are run on uppmax!

</div>

<div class="fragment">
<h3>Examples</h3>

All examples are available in lecture repository
<https://github.com/NBISweden/workshop-snakemake-byoc> under
`lectures/smkprofile`

<div>


# Cluster execution

From <https://snakemake.readthedocs.io/en/stable/executing/cluster.html>:

> Snakemake can make use of cluster engines that support shell scripts
> and have access to a common filesystem, (e.g. the Sun Grid Engine). In
> this case, Snakemake simply needs to be given a submit command that
> accepts a shell script as first positional argument:

```{r  byoc-2020-cluster-exec-example, engine='bash' }
snakemake --cluster qsub -j 32 
```

&nbsp;


## Important cluster command options

-   **&#x2013;cluster CMD:** execute snakemake rule with submit command; can be
    decorated e.g. `$ snakemake --cluster 'sbatch -c {threads}'`
-   **&#x2013;jobscript SCRIPT:** Provide a custom job script for submission to
    the cluster
-   **&#x2013;cluster-status STATUS:** status command for cluster execution to
    determine if job has finished


# Cluster execution example


## Snakefile

```{r  byoc-2020-slurm-snakefile, code=readLines('examples/1/Snakefile'), engine='python' }
 
```


## Slurm cluster execution

```{r  byoc-2020-slurm-exec-1, engine='bash' }
module load snakemake
snakemake --cluster "sbatch -A $ACCOUNT -p devel -n 1 -t 00:01:00" sleep_10.txt -j 1  -d examples/1 -s examples/1/Snakefile -F & 
```


## squeue

```{r  byoc-2020-slurm-exec-1-squeue, engine='bash' }
squeue -u $USER -A $ACCOUNT -l 
```

```
       JOBID PARTITION     NAME     USER    STATE       TIME TIME_LIMI  NODES NODELIST(REASON)
    15816917     devel snakejob    perun  PENDING       0:00      1:00      1 (Resources)


```


# Modifying the jobscript


## The jobscript

Snakemake wraps the rule in a `jobscript` which by default is located
at `/path/to/snakemake_installation/jobscript.sh`:
<small>

```{r  byoc-2020-show-snakemake-install-path, eval=TRUE, comment=NA, engine='bash' }
SMKPATH=$(python -c "import os; import snakemake; print(os.path.dirname(snakemake.__file__))")
## NB: for newer snakemake versions the path is $SMKPATH/executors/jobscript.sh
cat $SMKPATH/jobscript.sh 
```

</small>

A custom jobscript can be provided with the `--jobscript` option.


## An active jobscript

Generated jobscripts are output to a temporary directory in
`.snakemake` and can be accessed if the job is active or has failed:
<small>

```{r  byoc-2020-cat-jobscript, engine='bash' }
cat  examples/1/.snakemake/tmp.oihhb07u/snakejob.sleep.0.sh 
```

<div style="width:1200px;">

```
    #!/bin/sh
    # properties = {"type": "single", "rule": "sleep", "local": false, "input": [], "output": ["sleep_10.txt"], "wildcards": {"sec": "10"}, "params": {}, "log": [], "threads": 1, "resources": {}, "jobid": 0, "cluster": {}}
     cd /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/1 && \
    /sw/comp/python/3.7.2_rackham/bin/python3.7 \
    -m snakemake sleep_10.txt --snakefile /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/1/Snakefile \
    --force -j --keep-target-files --keep-remote \
    --wait-for-files /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/1/.snakemake/tmp.oihhb07u --latency-wait 5 \
     --attempt 1 --force-use-threads \
    --wrapper-prefix https://github.com/snakemake/snakemake-wrappers/raw/ \
    --directory /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/1   --allowed-rules sleep --nocolor --notemp --no-hooks --nolock \
    --mode 2  && touch /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/1/.snakemake/tmp.oihhb07u/0.jobfinished || (touch /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/1/.snakemake/tmp.oihhb07u/0.jobfailed; exit 1)


```

</div>
</small>


# Cluster status and missing errors


## Motivation: some errors go undetected

```{r  byoc-2020-timeout-example, engine='bash' }
snakemake -d examples/1 -s examples/1/Snakefile --cluster "sbatch -A $ACCOUNT -p devel -n 1 -t 00:01:00" sleep_90.txt -j 1 -F 
```

```
    Building DAG of jobs...
    Using shell: /usr/bin/bash
    Provided cluster nodes: 1
    Job counts:
    	count	jobs
    	1	sleep_90
    	1
    
    [Mon Oct  5 17:11:22 2020]
    rule sleep:
        output: sleep_90.txt
        jobid: 0
    
    Submitted job 0 with external jobid 'Submitted batch job 15817155'.
      C-c C-cTerminating processes on user request, this might take some time.
    Will exit after finishing currently running jobs.


```

Snakemake never returns from status page since it cannot determine
whether job completed successfully.


# Checking cluster status with custom script


## Custom status script

```{r  byoc-2020-cluster-status-script, code=readLines("examples/1/cluster-status.py"), engine='python' }
 
```


## Results

<small>

```{r  byoc-2020-timeout-example-with-status, engine='bash' }
snakemake -d examples/1 -s examples/1/Snakefile --cluster "sbatch -A $ACCOUNT -p devel -n 1 -t 00:01:00 --parsable" sleep_90.txt -j 1 -F --cluster-status ./cluster-status.py 
```

<div style="width:1200px;">

```
    Building DAG of jobs...
    <------
    ....
    ------->
    Submitted job 0 with external jobid '15817459'.
    [Mon Oct  5 19:07:35 2020]
    Error in rule sleep:
        jobid: 0
        output: sleep_90.txt
        shell:
            for i in $(seq 1 90); do echo $i >> sleep_90.txt; sleep 1; done
            (one of the commands exited with non-zero exit code; note that snakemake uses bash strict mode!)
        cluster_jobid: 15817459
    
    Error executing rule sleep on cluster (jobid: 0, external: 15817459, jobscript: /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/1/.snakemake/tmp.9od4w1lk/snakejob.sleep.0.sh). For error details see the cluster log and the log files of the involved rule(s).
    Shutting down, this might take some time.
    Exiting because a job execution failed. Look above for error message
    Complete log: /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/1/.snakemake/log/2020-10-05T190533.034527.snakemake.log


```

</div>
</small>


# Using a custom job submissions script

-   upon submission, snakemake formats a string `{submitcmd} "{jobscript}"` where
    `submitcmd` is constructed from the `--cluster` argument and `jobscript` from `--jobscript`
-   until now we have used `sbatch` as argument to `--cluster`


## A custom job submission script

```{r  byoc-2020-custom-job-submission-script, code=readLines("examples/1/submit-script.py"), engine='python' }
 
```


## Results

```{r  byoc-2020-custom-submit, engine='bash' }
snakemake -d examples/1 -s examples/1/Snakefile --cluster ./submit-script.py sleep_10.txt -j 1 -F --cluster-status ./cluster-status.py
## Grab job id and run
sacct -u $USER --format State,JobID,JobName,AllocCPUS -j 15817959 
```

<div style="width:1200px;">

```
         State        JobID    JobName  AllocCPUS 
    ---------- ------------ ---------- ---------- 
     COMPLETED 15817959     snakejob.+          2 
     COMPLETED 15817959.ba+      batch          2 
     COMPLETED 15817959.ex+     extern          2 


```

</div>


# SnakemakeProfiles


# slurm profile

<https://github.com/Snakemake-Profiles/slurm>


## Setting up

```
    $ cd examples
    $ cookiecutter https://github.com/Snakemake-Profiles/slurm.git
    You've downloaded /home/perun/.cookiecutters/slurm before. Is it okay to delete and re-download it? [yes]: profile_name [slurm]: 
    sbatch_defaults []: --account=$ACCOUNT --output=logs/slurm-%j.out --error=logs/slurm-%j.err
    cluster_config []: ../cluster-config.yaml
    Select advanced_argument_conversion:
    1 - no
    2 - yes
    Choose from 1, 2 [1]: 2


```

```{r  byoc-2020-slurm-profile, eval=TRUE, engine='bash' }
ls -l examples/slurm 
```


# Configuring jobs with cluster configuration


## Cluster configuration

Configure jobs with `--cluster-config`

```{r  byoc-2020-cluster-config, echo=FALSE, eval=TRUE,  comment=NA, class.source="sourceCode", class.output="sourceCode", engine='bash' }
cat examples/cluster-config.yaml 
```


## Running

```{r  byoc-2020-snakemake-profile-run, engine='bash' }
snakemake -d examples/1 -s examples/1/Snakefile --profile examples/slurm sleep_10.txt -j 10 -F
sacct -u $USER --format State,JobID,JobName,AllocCPUS -j 15818130 
```

```
         State        JobID    JobName  AllocCPUS 
    ---------- ------------ ---------- ---------- 
       PENDING 15818130     snakejob.+          3 


```


## BEWARE! cluster configuration is deprecated!

<small>

```
    --cluster-config FILE, -u FILE
                            A JSON or YAML file that defines the wildcards used in 'cluster'for specific rules, instead of having them specified in the Snakefile. For example, for rule 'job' you may define: { 'job' : { 'time' :
                            '24:00:00' } } to specify the time for rule 'job'. You can specify more than one file. The configuration files are merged with later values overriding earlier ones. This option is deprecated in favor of
                            using --profile, see docs. (default: [])


```

</small>


# Solving timeout with shadow rules


## Timeout issue again

Unfortunately snakemake does not remove incomplete output sleep\_90.txt
(bug? dunno).

```{r  byoc-2020-profile-timeout, engine='bash' }
snakemake -d examples/1 -s examples/1/Snakefile --profile examples/slurm sleep_90.txt -j 10 
```

```
    Building DAG of jobs...
    Nothing to be done.


```


## Workaround: shadow rule

With the `shadow` directive the rule runs in executed temporary
directories (default `.snakemake/shadow`)
<small>

```{r  byoc-2020-profile-shadow, engine='bash' }
snakemake -d examples/shadow -s examples/shadow/Snakefile --profile examples/slurm shadow_sleep_90.txt -j 10 
```

```
    Building DAG of jobs...
    Using shell: /usr/bin/bash
    Provided cluster nodes: 10
    Job counts:
    	count	jobs
    	1	sleep
    	1
    
    [Mon Oct  5 23:36:39 2020]
    rule sleep:
        output: shadow_sleep_90.txt
        jobid: 0
        wildcards: sec=90
    
    Submitted job 0 with external jobid '15818142'.
    [Mon Oct  5 23:38:41 2020]
    Error in rule sleep:
        jobid: 0
        output: shadow_sleep_90.txt
        shell:
            for i in $(seq 1 90); do echo $i >> shadow_sleep_90.txt; sleep 1; done
            (one of the commands exited with non-zero exit code; note that snakemake uses bash strict mode!)
        cluster_jobid: 15818142
    
    Error executing rule sleep on cluster (jobid: 0, external: 15818142, jobscript: /domus/h1/perun/teaching/workshop-snakemake-byoc/lectures/smkprofile/examples/shadow/.snakemake/tmp.lir98c26/snakejob.sleep.0.sh). For error details see the cluster log and the log files of the involved rule(s).
    Trying to restart job 0.
    
    [Mon Oct  5 23:38:42 2020]
    rule sleep:
        output: shadow_sleep_90.txt
        jobid: 0
        wildcards: sec=90
    
    Submitted job 0 with external jobid '15818144'.
    ...


```

</small>


# Adjusting jobs

Show example where attempt is used to increase time together with
jsonschema configuration


# Questions?

