---
title: Best practices in detail
subtitle: An overview of best practices, wrappers, schemas, report, config files, and more
author: Per Unneberg
date: "`r format(Sys.time(), '%d %B, %Y')`"
institute: NBIS
from: markdown+emoji
format:
  revealjs:
    theme:
      - white
      - ../custom.scss
#    css: ../revealjs.css
    self-contained: false
    toc: true
    toc-depth: 1
    slide-level: 2
    slide-number: true
    preview-links: true
    chalkboard: true
    # Multiple logos not possible; would need to make custom logo combining both logos
    footer: Snakemake BYOC 2022
    logo: https://nbis.se/assets/img/logos/nbislogo-green.svg
    smaller: true
    highlight-style: gruvbox
    fig-height: 3
    fig-width: 3
execute:
  echo: true
  warning: false
  cache: false
  include: true
  autodep: true
  eval: true
  error: true
knitr:
  opts_chunk:
    code-fold: false
    tidy: false
    fig-format: svg
---

## Setup  {.unnumbered .unlisted}



```{r libs }
#| echo: false
#| eval: true
#| cache: false
library(ggplot2)
library(viridis)
bw <- theme_bw(base_size=24) %+replace% theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
theme_set(bw)
snakemake_version <- system("snakemake --version", intern=TRUE)
```

- Examples based on more advanced snakefiles and code to run
- All snakefiles and code are available in code repository [https://github.com/NBISweden/snakemake_best_practice/](https://github.com/NBISweden/snakemake_best_practice/)
- Code has been run with Snakemake version `r snakemake_version`

# Snakemake best practices


## [Snakemake best practices summary](https://snakemake.readthedocs.io/en/stable/snakefiles/best_practices.html)

::: {.incremental}

Linting
: Snakemake (>=5.11) comes with a code quality checker (a so called
  linter). It is highly recommended to run the linter before
  publishing any workflow, asking questions on Stack Overflow or
  filing issues on Github.

Formatting
: There is an automatic formatter for Snakemake workflows, called
  Snakefmt, which should be applied to any Snakemake workflow before
  publishing it.

Testing
: It is a good idea to add some minimal test data and configure Github
  Actions for continuously testing the workflow on each new commit.

Structure
: Stick to a standardized structure.

Configuration
: Configuration of a workflow should be handled via config files and,
  if needed, tabular configuration like sample sheets (either via
  Pandas or PEPs). Use such configuration for metadata and experiement
  information, **not for runtime specific configuration like threads,
  resources and output folders**. For those, just rely on Snakemake’s
  CLI arguments like --set-threads, --set-resources,
  --set-default-resources, and --directory.

Filenames
: Try to keep filenames short, but informative.

Rules and functions
: Try to keep Python code like helper functions separate from rules.

Wrappers
: Make use of Snakemake wrappers whenever possible

:::

## A best practice repo

Clone the repo (`git clone
git@github.com:NBISweden/snakemake_best_practice.git`) and list
contents:

```{bash }
#| label: snakemake-byoc-2021-bp-overview
#| cache: true
#| eval: false
tree -a -L 2 -I '.snakemake|.git' snakemake_best_practice
```

```{bash }
#| label: snakemake-byoc-2021-bp-overview-run
#| cache: true
#| eval: true
#| echo: false
#| class-output: "scroll-1000"
tree -a -L 2 -I '.snakemake|.git|.ipynb_checkpoints|.myprofile|logs|reports|results|interim|*.~undo-tree~|*.png|*.zip|*.html' ../../snakemake_best_practice | sed "s/\.\.\/\.\.\///"
```

## What does it do?

Excerpts from README.md:
```{r code=readLines("../README.md")[c(1:25)]}
#| label: snakemake-byoc-2021-bp-readme
#| eval: false
#| highlight: false
```

::: {.fragment}

```{r code=readLines("../README.md")[c(106:112)]}
#| label: snakemake-byoc-2021-bp-readme-tail
#| eval: false
#| highlight: false
#| attr-source: startFrom="106"
```


:::

::: {.fragment}

Use a test data set for test driven development of the workflow. It
also gives a new user a quick idea of how to organize input files and
configuration.


:::




## Dry-run the test suite

```{bash }
#| label: snakemake-byoc-2021-dry-run
snakemake -s ../workflow/Snakefile -n -q -F
```

## Draw the workflow

```{bash }
#| label: dry-run-fig-command
#| echo: true
#| eval: false
snakemake -s ../workflow/Snakefile --rulegraph | dot | display
```
		



```{bash }
#| label: dry-run-fig
#| fig-format: svg
#| output: asis
#| echo: false
snakemake -s ../workflow/Snakefile --rulegraph | dot -T svg | grep -v "<!DOCTYPE" | grep -v "dtd"
```

## Snakefile

```{python code=readLines("../workflow/Snakefile")} 
#| label: Snakefile
#| eval: false
```

::: {.notes}

- explain pseudo-targets 
- point out the two common idioms for collecting targets:
1. expand
2. input functions

:::


## Stuff common to all snakefiles

<div style="font-size:80%">

`workflow/rules/common.smk`
```{python code=readLines("../workflow/rules/common.smk")}
#| label: common-smk
#| eval: false
```
</div>


## Input functions

`workflow/rules/inputfunctions.smk`
```{python code=readLines("../workflow/rules/inputfunctions.smk")[1:39]}
#| label: inputfunctions-smk
#| eval: false
```

## Pseudo-target functions


`workflow/rules/inputfunctions.smk`
```{python code=readLines("../workflow/rules/inputfunctions.smk")[41:76]}
#| label: inputfunctions-smk-2
#| eval: false
#| attr-source: startFrom="41"
```



## Rules

`workflow/rules/mapping.smk`
```{python code=readLines("../workflow/rules/mapping.smk")[4:31]}
#| label: mapping-smk-rule-example
#| eval: false
#| attr-source: startFrom="4"
```


# Scripts and wrappers

`script` directive:

- point to external script
- path relative to **file containing the directive**

## R script

`workflow/rules/qc.smk`:

```{python qc_plot_samtools_coverage, code=readLines("snakemake_best_practice/workflow/rules/qc.smk")[63:81], attr.source='startFrom="63"', eval=FALSE }

```

`workflow/scripts/plot_coverage.R`

```{r snakemake-byoc-2021-r-script, code=readLines("snakemake_best_practice/workflow/scripts/plot_coverage.R"), eval=FALSE }

```
S4 attributes map to rule directives (e.g. `snakemake@input[["png"]]`)

## rmarkdown scripts

`workflow/rules/qc.smk`:

```{python qc_rmarkdown, code=readLines("snakemake_best_practice/workflow/rules/qc.smk")[118:129], attr.source='startFrom="118"', eval=FALSE }

```
<div style="font-size:80%;">
`workflow/scripts/rmarkdown.Rmd`

```{r rmarkdown-script, code=readLines("snakemake_best_practice/workflow/scripts/rmarkdown.Rmd"), eval=FALSE }

```
</div>



## python script


`workflow/rules/qc.smk`:

<div style="font-size:80%;">
```{python qc_plot_samtools_coverage_collate, code=readLines("snakemake_best_practice/workflow/rules/qc.smk")[84:99], attr.source='startFrom="84"', eval=FALSE }

```
</div>

`workflow/scripts/plot_coverage.py`

<div style="font-size:80%;">

```{python snakemake-byoc-2021-python-script, code=readLines("snakemake_best_practice/workflow/scripts/plot_coverage.py"), eval=FALSE}

```
</div>

Rule directives accessible via `snakemake` object (e.g. `snakemake.input.txt`)


## jupyter notebook integration

`workflow/rules/qc.smk`

```{python jupyter-notebook-rule, code=readLines("snakemake_best_practice/workflow/rules/qc.smk")[102:115], eval=FALSE, attr.source='startFrom="102"'}

```
Generate output in `.tests` with 
```{bash jupyter-notebook-output, eval=FALSE}
snakemake --use-conda -s ../workflow/Snakefile reports/qc/notebook.html -j 1
```

<div class="fragment">

To edit, start `jupyter-notebook` and open `workflow/notebooks/notebook.py.ipynb`:

```{bash start-jupyter-notebook, eval=FALSE }
jupyter-notebook workflow/notebooks/notebook.py.ipynb
```

</div>

## Wrappers

`wrapper` directive: 

- Reusable wrapper scripts around e.g. command-line tools
- [The Snakemake Wrappers
  repository](https://snakemake-wrappers.readthedocs.io/en/stable/)
  contains a collection of reusable wrappers.
- accession format: `{version}/bio/{tool}/{command}`

<h3>Example</h3>

```{python qc-fastqc-wrapper, code=readLines("snakemake_best_practice/workflow/rules/qc.smk")[22:43], eval=FALSE, attr.source='startFrom="22"'}

```
[fastqc wrapper documentation](https://snakemake-wrappers.readthedocs.io/en/stable/wrappers/fastqc.html)

# Configuration and schemas

The workflow can be configured with a configuration file:

`workflow/config/config.yaml`

```{r configuration-example, code=readLines("snakemake_best_practice/config/config.yaml"), eval=FALSE, highlight=FALSE}

```

<div class="fragment">

`workflow/config/sample.tsv`

```{r configuration-example-sample, code=readLines("snakemake_best_practice/config/samples.tsv"), eval=FALSE, highlight=FALSE}

```

`workflow/config/reads.tsv`

```{r configuration-example-reads, code=readLines("snakemake_best_practice/config/reads.tsv")[1:4], eval=FALSE, highlight=FALSE}

```

</div>

<div class="fragment">

Question: is there a way to validate configuration files, require
inputs and make sure they conform to some predefined format?

</div>


## Configuration schemas

Schema benefits according to [https://json-schema.org/]():

- describes your existing data formats
- provides human- and machine-readable **documentation**
- validates data input

<div class="fragment">

`workflow/schemas/samples.schema.yaml`

```{r configuration-sample-schema, code=readLines("snakemake_best_practice/workflow/schemas/samples.schema.yaml"), eval=FALSE, highlight=FALSE}

```
</div>


## Configuration schemas

<div style="font-size:65%">

`workflow/schemas/config.schema.yaml`

```{r configuration-config-schema, code=readLines("snakemake_best_practice/workflow/schemas/config.schema.yaml"), eval=FALSE, highlight=FALSE}

```

<div class="fragment">

Recall validation step in `workflow/rules/common.smk`:

```{python configuration-config-schema-validation, code=readLines("snakemake_best_practice/workflow/rules/common.smk")[15:20], eval=FALSE, attr.source='startFrom="15"'}

```

</div>
</div>


# Reports


From snakemake 5.1 and on, generate detailed self-contained HTML
reports that encompass runtime statistics, provenance information,
workflow topology and results

## The report directive

`workflow/Snakefile`:

```{python snakemake-report, code=readLines("snakemake_best_practice/workflow/Snakefile")[13:13], eval=FALSE, attr.source='startFrom="13"'}

```

Workflow report template defined by `workflow/report/workflow.rst`.


<div class="fragment">

Use `report` flag to target results for inclusion in report, which
could optionally point to an rst file for captioning.

`workflow/rules/qc.smk`:

```{python r-plot-report, code=readLines("snakemake_best_practice/workflow/rules/qc.smk")[63:81], eval=FALSE, attr.source='startFrom="63"'}

```
</div>

## Workflow rst template files

Template files are [restructured text
format](https://docutils.sourceforge.io/docs/user/rst/quickstart.html)
(rst) files.

`workflow/report/workflow.rst`:

```{r snakemake-workflow-rst, code=readLines("snakemake_best_practice/workflow/report/workflow.rst"), eval=FALSE, highlight=FALSE }

```
`workflow/report/coverage.rst`:

```{r snakemake-coverage-rst, code=readLines("snakemake_best_practice/workflow/report/coverage.rst"), eval=FALSE, highlight=FALSE }

```

<div class="fragment">
<h3>Creating the report</h3>

```{r cd_to_sbp_report, echo=FALSE }
knitr::opts_knit$set(root.dir="snakemake_best_practice/.test")
```

```{bash snakemake-create-report, eval=TRUE, cache=TRUE }
snakemake -s ../workflow/Snakefile --report report.html
```

```{r cd_back_sbp_report, echo=FALSE }
knitr::opts_knit$set(root.dir=curdir)
```

</div>

# Coding practices and hints

<h3>snakemake --lint</h3>

A linter is a code quality checker that analyzes your code and
highlights issues that need to be resolved to follow best practices.


```{r cd_to_snakemake_best_practice_2, echo=FALSE }
knitr::opts_knit$set(root.dir="snakemake_best_practice")
```


```{bash snakemake-lint }
snakemake --lint
```


<h3>snakefmt</h3>

[snakefmt](https://github.com/snakemake/snakefmt) is an automated code
formatter that should be applied to the workflow prior to publication.

```{bash snakemake-fmt }
snakefmt --compact-diff workflow/Snakefile
```

```{r cd_back_2, echo=FALSE }
knitr::opts_knit$set(root.dir=curdir)
```


## Pre-commit - for the git power user

[Git hooks](https://git-scm.com/docs/githooks) can
be used to identify simple issues before submission to code review.

[Pre-commit](https://pre-commit.com) is
a "framework for managing and maintaining multi-language pre-commit
hooks".

<h3>Write a config file</h3>

<div style="font-size:80%;">

`.pre-commit-config.yaml`:

```{r pre-commit-config, code=readLines("snakemake_best_practice/.pre-commit-config.yaml"), eval=FALSE, highlight=FALSE }

```

</div>

<h3>Usage</h3>

Install git hooks

```{bash pre-commit, eval=FALSE}
pre-commit install
```

and see how many warnings you get when you try to commit!

## Github actions for continuous integration

[Snakemake github
action](https://github.com/snakemake/snakemake-github-action) allows
running the test suite on github to make sure commits and pull
requests don't break the workflow.

`.github/workflows/main.yaml`:

<div style="font-size:80%;">

```{r github-actions, code=readLines("snakemake_best_practice/.github/workflows/main.yaml"), eval=FALSE, highlight=FALSE }

```

</div>

## On project file structure vs workflow file structure

Example from my config which is loosely modelled on the
[drivendata](http://drivendata.github.io/cookiecutter-data-science/)
setup and similar to the NBIS reproducibility file structure:

```{bash project-file-structure, cache=TRUE, echo=FALSE }
tree -a -d -L 2 -I '.snakemake|.git' project
```

Different snakemake workflows live in `opt` (see [File System Hierachy
standard](https://www.pathname.com/fhs/pub/fhs-2.3.html) for choice of
name). Launching from project root could then look like

```{bash project-structure-launch, eval=FALSE}
snakemake -s opt/datasources-smk/workflow/Snakefile -j 1
```



# Questions?
